---
title: "Backfits"
author: "Jason Willwerscheid"
date: "8/17/2019"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Although I've done much to speed up backfits, they remain slow relative to greedy fits. The question, then, is whether they're worth the wait. Here, I'll argue that while the quantitative improvements are relatively small --- one can obtain a similar increase in ELBO in a fraction of the time by greedily adding a handful of additional factors --- the qualitative improvements can be dramatic.

In this analysis, I take my preferred fit from a previous analysis of [variance structures](var_type.html) (the fit that "pre-scales" cells) and compare it against a backfitted version of the same. The code used to produce the fits can be viewed [here](https://github.com/willwerscheid/scFLASH/blob/master/code/backfit/backfit_fits.R).

```{r load.res}
source("./code/utils.R")
droplet <- readRDS("./data/droplet.rds")
droplet <- preprocess.droplet(droplet)
res <- readRDS("./output/backfit/backfit_fits.rds")

progress.g <- data.table::fread("./output/backfit/greedy_output.txt")
progress.b <- data.table::fread("./output/backfit/backfit_output.txt")
```

## Results: ELBO

As a quantitative measure of how much backfitting improves the fit, I plot the ELBO attained after each of the last eight greedy factors have been added and after each of the first 60 backfitting iterations (backfitting continues for a total of 100 iterations, but the ELBO remains mostly level after the first 50 or so iterations).

From the perspective of the ELBO, backfitting offers about as much improvement as the greedy addition of 5-6 factors. However, a single backfitting iteration takes almost 30 seconds, whereas the entire greedy fit can be performed in less than two minutes. If the prospect of an hours-long backfit is too daunting, then a partial backfit might be sufficient: most of the improvement occurs within the first ten backfitting iterations.

```{r progress}
g.iter <- nrow(progress.g)
b.iter <- nrow(progress.b)
progress.g$elapsed.time <- (res$greedy$elapsed.time * 1:g.iter / g.iter) / 60
progress.b$elapsed.time <- (res$greedy$elapsed.time 
                            + res$backfit$elapsed.time * 1:b.iter / b.iter) / 60

# Only the final iteration for each greedy factor is needed.
progress.g <- subset(progress.g, c((progress.g$Factor[1:(nrow(progress.g) - 1)]
                                    != progress.g$Factor[2:nrow(progress.g)]), TRUE))

progress.df <- rbind(progress.g, progress.b)

ggplot(subset(progress.df, !(Factor %in% as.character(1:12)) & Iter < 61), 
       aes(x = elapsed.time, y = Obj, color = Type)) +
  geom_point() +
  labs(x = "elapsed time (min)", y = "ELBO (unadjusted)",
       title = "Greedy factors 13-20 and first 60 backfitting iterations")
```

## Results: Factor comparisons

The ELBO does not tell the whole story. Plotting the factors reveals clear improvements in the backfitted factors. Most obviously, factors 3, 6, 8, 11, 16, and 17 are much sparser and are more easily readable as cell-type specific factors. In addition, factor 2 more clearly indexes a basal-to-club gradation without ciliated cells mucking up the situation.

```{r factors}
plot.factors(res$greedy, droplet$cell.type, kset = 1:20, title = "Greedy")
plot.factors(res$backfit, droplet$cell.type, kset = 1:20, title = "Backfit")
```

Factors above the dashed line in the plot below are factors that have increased in importance, while factors below the line have diminished in importance. Most interestingly, backfitting uncovers a large club-cell specific factor that is relatively unimportant in the greedy fit (factor 19).

```{r pve}
pve.df <- data.frame(k = 1:res$greedy$fl$n.factors, 
                     greedy = res$greedy$fl$pve, 
                     backfit = res$backfit$fl$pve)
ggplot(pve.df, aes(x = greedy, y = backfit)) +
  geom_text(label = pve.df$k, size = 3) + 
  geom_abline(slope = 1, linetype = "dashed") + 
  scale_x_log10() + scale_y_log10() +
  labs(title = "Proportion of variance explained")
```

