---
title: "Deng et al. dataset"
author: "Jason Willwerscheid"
date: "3/2/2022"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

## Overview

```{r read.data}
library(tidyverse)
library(flashier)
library(ggrepel)
library(singleCellRNASeqMouseDeng2014)

counts <- exprs(Deng2014MouseESC)
meta_data <- pData(Deng2014MouseESC)
gene_names <- rownames(counts)

preprocess <- function(dat, min.nzcts = 10) {
  size.factors <- colSums(dat)
  size.factors <- size.factors / mean(size.factors)
  gene_cts <- rowSums(dat > 0)
  dat <- dat[gene_cts >= min.nzcts, ]

  lunpc <- max(1 / min(size.factors) - 1 / max(size.factors), 1)
  fl.dat <- log1p(t(t(dat) / size.factors) / lunpc)

  return(list(
    dat = dat,
    fl.dat = fl.dat,
    size.factors = size.factors,
    excluded.genes = gene_cts < min.nzcts)
  )
}
Deng <- preprocess(counts)
```

The dataset, from [Deng et al.](https://www.science.org/doi/10.1126/science.1245316), is made available by `kkdey`'s `R` package `singleCellRNASeqMouseDeng2014`, which I installed using command `remotes::install_github("kkdey/singleCellRNASeqMouseDeng2014")`.

After removing genes with nonzero counts in fewer than 10 cells, there remain counts for 17176 genes and 259 cells. Each cell has been labelled as one of 10 cell types (or rather, one of 10 embryonic stages ranging from zygote to late blastocyte).

I fit 6, 10, and 25 semi-nonnegative EBMF factors using `flashier`. Code is [here](https://github.com/willwerscheid/scFLASH/blob/master/code/deng/deng.R). In the heatmaps below, rows correspond to individual cells, and factors (columns) are arranged in order of decreasing proportion of variance explained.

```{r fl.plot}
do.heatmap <- function(fl) {
  FF <- ldf(fl, type = "I")$F
  FF <- FF[, order(fl$pve, decreasing = TRUE)]
  colnames(FF) <- 1:ncol(FF) 
  
  cell_type <- meta_data$cell_type
  
  tib <- as_tibble(FF) %>%
    mutate(Cell.idx = row_number()) %>%
    mutate(Cell.type = cell_type) %>%
    mutate(Cell.type = fct_relevel(Cell.type, c(
      "zy",
      "early2cell", "mid2cell", "late2cell",
      "4cell", "8cell", "16cell",
      "earlyblast", "midblast", "lateblast"
    )))
  tib <- tib %>%
    pivot_longer(
      -c(Cell.idx, Cell.type),
      names_to = "Factor",
      values_to = "Loading",
      values_drop_na = TRUE
    ) %>%
    mutate(Factor = as.numeric(Factor))
  
  cell_type_breaks <- c(1, which(cell_type[2:nrow(tib)] != cell_type[1:(nrow(tib) - 1)]))
  ggplot(tib, aes(x = Factor, y = -Cell.idx, col = Loading)) +
    geom_tile() +
    scale_color_gradient(low = "black", high = "red") +
    labs(y = "") +
    scale_y_continuous(breaks = -cell_type_breaks,
                       labels = levels(tib$Cell.type))
}
```

### 6-factor fit

```{r fl6}
fl6 <- readRDS("./output/deng/deng_fl6.rds")
do.heatmap(fl6)
```

### 10-factor fit

```{r fl10}
fl10 <- readRDS("./output/deng/deng_fl10.rds")
do.heatmap(fl10)
```

### 25-factor fit

```{r fl25}
fl25 <- readRDS("./output/deng/deng_fl25.rds")
do.heatmap(fl25)
```

## Top genes

Many of the factors in the 25-factor fit seem to primarily capture noise in individual cells. I show volcano plots for the more interesting factors, which I've arranged in rough ontogenetic order. I label the top 20 genes by (absolute) z-score (defined as posterior mean / posterior SD) as well as the top 20 by (absolute) posterior mean.

```{r volcano.code}
do.volcano.plot <- function(fl, k) {
  k <- order(fl$pve, decreasing = TRUE)[k]
  tib <- tibble(
    pm = fl$L.pm[, k],
    z = abs(fl$L.pm[, k]) / pmax(sqrt(.Machine$double.eps), fl$L.psd[, k]),
    exprmean = log10(rowMeans(Deng$dat)),
    SYMBOL = rownames(fl$L.pm)
  ) %>%
    mutate(SYMBOL = ifelse(
      z > sort(z, decreasing = TRUE)[21] |
        abs(pm) > sort(abs(pm), decreasing = TRUE)[21], SYMBOL, ""))

  plt <- ggplot(tib, aes(x = pm, y = z, color = exprmean, label = SYMBOL)) +
    geom_point() +
    scale_color_gradient2(low = "deepskyblue", mid = "gold", high = "orangered",
                          na.value = "gainsboro",
                          midpoint = mean(range(tib$exprmean))) +
    scale_y_sqrt() +
    geom_text_repel(color = "darkgray",size = 2.25, fontface = "italic",
                    segment.color = "darkgray", segment.size = 0.25,
                    min.segment.length = 0, na.rm = TRUE) +
    theme_minimal() +
    labs(
      x = "Factor Loading (posterior mean)",
      y = "|z-score|",
      color = "Mean Expression (log10)"
    ) +
    theme(legend.position = "bottom")

  return(plt)
}
```

### Factor 10: zygote to early 2-cell
```{r vp10}
do.volcano.plot(fl25, 10)
```

### Factor 2: zygote to mid 2-cell

Overlaps with the blue cluster in [Dey et al.](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1006599#sec008):
```{r vp2}
do.volcano.plot(fl25, 2)
```

### Factor 8: mid to late 2-cell

Overlaps with the magenta cluster in Dey et al.:
```{r vp8}
do.volcano.plot(fl25, 8)
```

### Factor 9: late 2-cell to 4-cell
```{r vp9}
do.volcano.plot(fl25, 9)
```

### Factor 3: late 2-cell to 16-cell
Some overlap with the yellow cluster in Dey et al.:
```{r vp3}
do.volcano.plot(fl25, 3)
```

### Factor 6: 16-cell to late blastocyte
Overlaps with the green cluster in Dey et al., but also the orange to some extent:
```{r vp6}
do.volcano.plot(fl25, 6)
```

### Factor 4: early to late blastocyte
Overlaps with the orange cluster in Dey et al.:
```{r vp4}
do.volcano.plot(fl25, 4)
```

### Factor 16: early blastocyte
```{r vp16}
do.volcano.plot(fl25, 16)
```

### Factor 5: early to mid blastocyte
Overlaps with the purple cluster in Dey et al.:
```{r vp5}
do.volcano.plot(fl25, 5)
```

### Factor 11: mid blastocyte
```{r vp11}
do.volcano.plot(fl25, 11)
```

### Factor 7: late blastocyte
```{r vp7}
do.volcano.plot(fl25, 7)
```
