---
title: "Deng et al. dataset: nonnegative fits"
author: "Jason Willwerscheid"
date: "3/28/2022"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

## Overview

```{r read.data}
library(tidyverse)
library(flashier)
library(singleCellRNASeqMouseDeng2014)
library(Rtsne)
library(ggrepel)
# library(fastTopics)

counts <- exprs(Deng2014MouseESC)
meta_data <- pData(Deng2014MouseESC)
gene_names <- rownames(counts)

preprocess <- function(dat, min.nzcts = 10) {
  size.factors <- colSums(dat)
  size.factors <- size.factors / mean(size.factors)
  gene_cts <- rowSums(dat > 0)
  dat <- dat[gene_cts >= min.nzcts, ]

  lunpc <- max(1 / min(size.factors) - 1 / max(size.factors), 1)
  fl.dat <- log1p(t(t(dat) / size.factors) / lunpc)

  return(list(
    dat = dat,
    fl.dat = fl.dat,
    size.factors = size.factors,
    excluded.genes = gene_cts < min.nzcts)
  )
}
Deng <- preprocess(counts)
```

I give semi-nonnegative and non-negative `flashier` fits to the Deng et al. dataset (see [here](deng.html) for an introduction). For both, I added factors greedily. I used point-exponential priors for factors and point-Laplace or point-exponential priors for loadings. I also fix a "mean factor" (shown as factor 1 below).

```{r snmf}
snmf <- readRDS("./output/deng/snmf.rds")
nmf <- readRDS("./output/deng/nmf.rds")

snmf$fl <- flash.reorder.factors(snmf$fl, c(1, order(snmf$fl$pve[-1], decreasing = TRUE) + 1))
nmf$fl <- flash.reorder.factors(nmf$fl, c(1, order(nmf$fl$pve[-1], decreasing = TRUE) + 1))

get.factors <- function(res, colnames.prefix) {
  FF <- ldf(res$fl, type = "I")$F
  colnames(FF) <- paste0(colnames.prefix, 1:ncol(FF))
  return(FF)
}

snmf.F <- get.factors(snmf, "SNMF")
nmf.F <- get.factors(nmf, "NMF")
tib <- cbind(snmf.F, nmf.F)

tsne_res <- Rtsne(
    tib,
    dims = 1,
    perplexity = pmax(1, floor((nrow(tib) - 1) / 3) - 1),
    pca = FALSE,
    normalize = FALSE,
    theta = 0.1,
    check_duplicates = FALSE,
    verbose = FALSE
  )$Y[, 1]

tib <- as_tibble(tib) %>%
  mutate(tsne_res = unlist(tsne_res)) %>%
  mutate(Cell.type = fct_relevel(meta_data$cell_type, c(
    "zy",
    "early2cell", "mid2cell", "late2cell",
    "4cell", "8cell", "16cell",
    "earlyblast", "midblast", "lateblast"
  )))

tsne.zy <- tib %>% filter(Cell.type == "zy") %>% summarize(mean(tsne_res))
tsne.late <- tib %>% filter(Cell.type == "lateblast") %>% summarize(mean(tsne_res))
if (tsne.zy < tsne.late) {
  tib <- tib %>%
    arrange(Cell.type, tsne_res)
} else {
  tib <- tib %>%
    arrange(Cell.type, -tsne_res)
}

tib <- tib %>%
  mutate(Cell.idx = row_number()) %>%
  select(-tsne_res)
  
tib <- tib %>%
  pivot_longer(
    -c(Cell.idx, Cell.type),
    names_to = "Factor",
    values_to = "Loading",
    values_drop_na = TRUE
  ) %>%
  mutate(
    Fit = factor(str_remove(Factor, "[0-9]+"), levels = c("SNMF", "NMF")),
    Factor = as.numeric(str_extract(Factor, "[0-9]+"))
  )

cell_type <- tib %>% group_by(Cell.idx) %>% summarize(Cell.type = Cell.type[1]) %>% pull(Cell.type)
cell_type_breaks <- c(-4, which(cell_type[2:nrow(tib)] != cell_type[1:(nrow(tib) - 1)]))
  
ggplot(tib, aes(x = Factor, y = -Cell.idx, fill = Loading)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  labs(x = "Factor", y = "") +
  scale_y_continuous(breaks = -cell_type_breaks,
                     minor_breaks = NULL,
                     labels = levels(tib$Cell.type)) +
  theme_minimal() +
  theme(axis.text.y=element_text(angle = 45, size = 6)) +
  facet_wrap(~Fit, nrow = 1)
```

## Factors

I've tried to group together factors that are very similar for ease of comparison. The headers roughly indicate the embryonic stage where the factors are active. 

```{r volcano.code}
do.volcano.plot <- function(nmf.k = 0, snmf.k = 0, plt.title = "") {
  make.tib <- function(fit, fl, k) {
    if (k == 0) {
      return(tibble())
    }
    
    tib <- tibble(
      fit = fit,
      pm = ldf(fl, type = "I")$L[, k],
      z = abs(fl$L.pm[, k]) / pmax(sqrt(.Machine$double.eps), fl$L.psd[, k]),
      exprmean = log10(rowMeans(Deng$dat)),
      SYMBOL = rownames(fl$L.pm)
    ) %>%
      mutate(SYMBOL = ifelse(
        z * (abs(pm) > 0.1) > sort(z * (abs(pm) > 0.1), decreasing = TRUE)[16] |
          (pm > 0.1 & pm > sort(pm, decreasing = TRUE)[11]) |
          (pm < -0.1 & pm < sort(pm, decreasing = FALSE)[11]), 
        SYMBOL, 
        ""
      )) %>%
      mutate(fit = paste(fit, "Factor", k))
    return(tib)
  }
  nmf.tib <- make.tib("NMF", nmf$fl, nmf.k)
  snmf.tib <- make.tib("SNMF", snmf$fl, snmf.k)
  tib <- nmf.tib %>% bind_rows(snmf.tib)
    
  plt <- ggplot(tib, aes(x = pm, y = z, color = exprmean, label = SYMBOL)) +
    geom_point() +
    scale_color_gradient2(low = "deepskyblue", mid = "gold", high = "orangered",
                          na.value = "gainsboro",
                          midpoint = mean(range(tib$exprmean))) +
    scale_y_sqrt() +
    geom_text_repel(color = "darkgray",size = 2.25, fontface = "italic",
                    segment.color = "darkgray", segment.size = 0.25,
                    min.segment.length = 0, na.rm = TRUE, max.overlaps = 20) +
    theme_minimal() +
    labs(
      x = "Factor Loading (posterior mean)",
      y = "|z-score|",
      color = "Mean Expression (log10)",
      title = plt.title
    ) +
    theme(legend.position = "bottom") +
    facet_wrap(~fit, scales = "free", ncol = 1)

  return(plt)
}
```

### Zygote to mid 2cell

```{r v1}
do.volcano.plot(0, 2)
do.volcano.plot(6, 0)
do.volcano.plot(2, 0, "(All but zygote to mid 2cell)")
do.volcano.plot(3, 0, "(All but zygote to mid 2cell)")
```

### Mid to late 2cell

```{r v2}
do.volcano.plot(0, 9)
do.volcano.plot(10, 0)
```

### Late 2cell / 4cell

```{r v3}
do.volcano.plot(0, 8)
do.volcano.plot(8, 0)
```

### 8cell / 16cell

```{r v4}
do.volcano.plot(0, 6)
do.volcano.plot(4, 0)
```

### Early blastocyte

```{r v5}
do.volcano.plot(0, 12)
do.volcano.plot(13, 0, "All but early blastocyte")
```

### Blastocytes

```{r v6}
do.volcano.plot(0, 3)
do.volcano.plot(0, 5)
do.volcano.plot(0, 4)
do.volcano.plot(7, 0)
do.volcano.plot(5, 0)
```

### All over

```{r v7}
do.volcano.plot(0, 10)
do.volcano.plot(9, 0)
```

### NMF-only factors

```{r v8}
do.volcano.plot(12, 0, "Zygote / Early 2cell")
do.volcano.plot(11, 0, "Late Blastocytes")
```
