---
title: "Deng et al. dataset: nonnegative fits"
author: "Jason Willwerscheid"
date: "3/28/2022"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

## Overview

```{r read.data}
library(tidyverse)
library(flashier)
library(ggrepel)
library(singleCellRNASeqMouseDeng2014)
library(Rtsne)
library(fastTopics)

counts <- exprs(Deng2014MouseESC)
meta_data <- pData(Deng2014MouseESC)
gene_names <- rownames(counts)

preprocess <- function(dat, min.nzcts = 10) {
  size.factors <- colSums(dat)
  size.factors <- size.factors / mean(size.factors)
  gene_cts <- rowSums(dat > 0)
  dat <- dat[gene_cts >= min.nzcts, ]

  lunpc <- max(1 / min(size.factors) - 1 / max(size.factors), 1)
  fl.dat <- log1p(t(t(dat) / size.factors) / lunpc)

  return(list(
    dat = dat,
    fl.dat = fl.dat,
    size.factors = size.factors,
    excluded.genes = gene_cts < min.nzcts)
  )
}
Deng <- preprocess(counts)
```

```{r fl.plot}
do.heatmap <- function(res) {
  fl <- res$fl
  
  FF <- ldf(fl, type = "I")$F
  FF <- FF[, -1]
  FF <- FF[, order(res$fl$pve[-1], decreasing = TRUE)]
  colnames(FF) <- 1:ncol(FF) 
  
  cell_type <- meta_data$cell_type
  
  tib <- as_tibble(FF) %>%
    mutate(Cell.type = cell_type) %>%
    mutate(Cell.type = fct_relevel(Cell.type, c(
      "zy",
      "early2cell", "mid2cell", "late2cell",
      "4cell", "8cell", "16cell",
      "earlyblast", "midblast", "lateblast"
    )))
  
  tsne_res <- Rtsne(
    as.matrix(tib %>% select(-Cell.type)),
    dims = 1,
    perplexity = pmax(1, floor((nrow(tib) - 1) / 3) - 1),
    pca = FALSE,
    normalize = FALSE,
    theta = 0.1,
    check_duplicates = FALSE,
    verbose = FALSE
  )$Y[, 1]
  tib <- tib %>%
    mutate(tsne_res = unlist(tsne_res)) %>%
    arrange(Cell.type, tsne_res) %>%
    mutate(Cell.idx = row_number()) %>%
    select(-tsne_res)
  
  tib <- tib %>%
    pivot_longer(
      -c(Cell.idx, Cell.type),
      names_to = "Factor",
      values_to = "Loading",
      values_drop_na = TRUE
    ) %>%
    mutate(Factor = as.numeric(Factor))
  
  cell_type_breaks <- c(1, which(cell_type[2:nrow(tib)] != cell_type[1:(nrow(tib) - 1)]))
  ggplot(tib, aes(x = Factor, y = -Cell.idx, fill = Loading)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "red") +
    labs(y = "") +
    scale_y_continuous(breaks = -cell_type_breaks,
                       minor_breaks = NULL,
                       labels = levels(tib$Cell.type)) +
    theme_minimal() +
    geom_hline(yintercept = -cell_type_breaks, size = 0.1)
}

do.struct.plot <- function(fl, kset, group_by_embryo = FALSE) {
  tm <- init_poisson_nmf(X = t(Deng$dat), k = fl$n.factors, init.method = "random")
  tm <- poisson2multinom(tm)
  L <- ldf(fl)$F %*% diag(ldf(fl)$D)
  L <- L[, order(fl$pve, decreasing = TRUE)]
  colnames(L) <- paste0("k", 1:ncol(L))
  L[, setdiff(1:fl$n.factors, kset)] <- 0
  tm$L <- L
  topic_colors <- rep("black", fl$n.factors)
  topic_colors[kset] <- c("gainsboro", "forestgreen", "tomato", "skyblue", "royalblue",
                          "darkorange", "peru", "gold", "limegreen", "darkmagenta")[1:length(kset)]
  cell_type <- factor(
    meta_data$cell_type,
    levels = c("zy", "early2cell", "mid2cell", "late2cell", "4cell", "8cell",
               "16cell", "earlyblast", "midblast", "lateblast")
  )
  embryo <- factor(
    paste0(meta_data$cell_type, "/", meta_data$embryo_id),
    levels = paste0(rep(levels(cell_type), each = length(levels(meta_data$embryo_id))),
                   "/", levels(meta_data$embryo_id))
  )
  embryo <- droplevels(embryo)
  embed_with_pca <- function (fit, ...) {
    drop(pca_from_topics(fit, dims = 1,...))
  }
  if (group_by_embryo) {
    grp <- embryo
  } else {
    grp <- cell_type
  }
  set.seed(1)
  structure_plot(tm, grouping = grp, colors = topic_colors,
                 gap = 2, topics = kset, embed_method = embed_with_pca) +
    labs(y = "loading", color = "factor", fill = "factor")
}
```

I give a semi-nonnegative and three non-negative `flashier` fits to the Deng et al. dataset (see [here](deng.html) for an introduction). The nonnegative fits were obtained by backfitting from a greedy nonnegative fit and from a NNMF obtained via `NNLM`.

### Semi-nonnegative fit

```{r smnf, results = 'asis'}
smnf <- readRDS("./output/deng/smnf.rds")
cat("ELBO:", smnf$fl$elbo, "\n")
units(smnf$t) <- "mins"
cat("Time to fit:", format(smnf$t, digits = 1), "\n")
cat("Maximum iterations reached:", !is.null(smnf$fl$flash.fit$maxiter.reached), "\n")
do.heatmap(smnf)
```

<!-- ### Greedy nonnegative fit -->

<!-- ```{r greedy2} -->
<!-- greedy2 <- readRDS("./output/deng/greedy2.rds") -->
<!-- do.heatmap(greedy2) -->
<!-- ``` -->

### Nonnegative fit (from greedy)

```{r bf2, results = 'asis'}
greedy <- readRDS("./output/deng/greedy.rds")
greedy2 <- readRDS("./output/deng/greedy2.rds")
bf <- readRDS("./output/deng/bf.rds")
bf2 <- readRDS("./output/deng/bf2.rds")
bf.t <- bf$t + bf2$t + greedy$t + greedy2$t
cat("ELBO:", bf2$fl$elbo, "\n")
units(bf.t) <- "mins"
cat("Time to fit:", format(bf.t, digits = 1), "\n")
cat("Maximum iterations reached:", !is.null(bf2$fl$flash.fit$maxiter.reached), "\n")
do.heatmap(bf2)
```

<!-- ### NNLM fit -->

<!-- ```{r nnlm} -->
<!-- nnlm <- readRDS("./output/deng/nnlm.rds") -->
<!-- do.heatmap(nnlm) -->
<!-- ``` -->

### Nonnegative fit (from NNLM)

```{r nnlmbf, results = 'asis'}
nnlmbf <- readRDS("./output/deng/nnlmbf.rds")
cat("ELBO:", nnlmbf$fl$elbo, "\n")
units(nnlmbf$t) <- "mins"
cat("Time to fit:", format(nnlmbf$t, digits = 1), "\n")
cat("Maximum iterations reached:", !is.null(nnlmbf$fl$flash.fit$maxiter.reached), "\n")
do.heatmap(nnlmbf)
```

### Nonnegative fit (from NNLM, shifted point-exponential priors)
```{r nzpe, results = 'asis'}
nzpe <- readRDS("./output/deng/nzpe.rds")
nzpe2 <- readRDS("./output/deng/nzpe2.rds")
# shifts <- sapply(nzpe$fl$F.ghat[-1], function(k) k$shift[1])
# nzpe$fl$flash.fit$EF[[2]][, -1] <- nzpe$fl$flash.fit$EF[[2]][, -1] - rep(shifts, each = ncol(Deng$fl.dat))
nzpe.t <- nzpe$t + nzpe2$t
cat("ELBO:", nzpe2$fl$elbo, "\n")
units(nzpe.t) <- "mins"
cat("Time to fit:", format(nzpe.t, digits = 1), "\n")
cat("Maximum iterations reached:", !is.null(nzpe2$fl$flash.fit$maxiter.reached), "\n")
do.heatmap(nzpe2)
```
