---
title: "Montoro et al. droplet dataset"
author: "Jason Willwerscheid"
date: "8/12/2019"
output:
  workflowr::wflow_html:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

```{r read.data}
droplet <- readRDS("./data/droplet.rds")
source("./code/utils.R")
```

The droplet-based 3' scRNA-seq dataset analyzed in Montoro et al. (2018) can be obtained [here](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE103354). It includes counts for 18,388 genes and 7,193 cells. Only 9.3% of all counts are nonzero. The data takes up 143 MB when loaded into memory as a sparse `Matrix` object.

The authors assign each cell to one of seven cell types. The cells are also labelled according to the mice they were taken from.

```{r cell.labels}
cell.type <- as.factor(sapply(strsplit(colnames(droplet), "_"), `[`, 3))
plot.category(cell.type, title = "Number of cells per cell type")
mouse <- as.factor(sapply(strsplit(colnames(droplet), "_"), `[`, 1))
plot.category(mouse, title = "Number of cells per mouse")
```

Library size is distributed as follows.

```{r cells}
plot.libsize(droplet)
```

Note the presence of extreme outliers. There are two of them, and both turn out to be goblet cells. I think that it would be a good idea to remove them from the dataset before analysis.

```{r cell.outliers}
table(cell.type[which(colSums(droplet) > 30000)])
```

Mean expression is distributed as follows.

```{r genes}
plot.meanexp(droplet)
```

There are two genes that are on average much more highly expressed than any other gene: Bpifa1 and Scgb1a1. Both have bimodal distributions.

```{r gene.outliers}
high.exp <- names(which(rowMeans(droplet) > 200))
for (gene in high.exp) {
  plot(plot.gene(droplet, gene))
}
```
