---
title: "PBMCs dataset: nonnegative fits"
author: "Jason Willwerscheid"
date: "3/30/2022"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

## Overview

```{r read.data}
library(tidyverse)
library(flashier)
library(ggrepel)
library(Matrix)
library(Rtsne)
library(fastTopics)

preprocess <- function(dat, min.nzcts = 10) {
  size.factors <- colSums(dat)
  size.factors <- size.factors / mean(size.factors)
  gene_cts <- rowSums(dat > 0)
  dat <- dat[gene_cts >= min.nzcts, ]

  lunpc <- max(1 / min(size.factors) - 1 / max(size.factors), 1)
  fl.dat <- log1p(t(t(dat) / size.factors) / lunpc)

  return(list(
    dat = dat,
    fl.dat = fl.dat,
    size.factors = size.factors,
    pc = lunpc,
    excluded.genes = gene_cts < min.nzcts)
  )
}

pbmc <- readRDS("../flashier-chapter/data/pbmc.rds")
pbmc <- pbmc[, colSums(pbmc) < 15000]
pbmc <- preprocess(pbmc)

pbmc.celltype <- sapply(strsplit(colnames(pbmc$fl.dat), "_"), `[[`, 2)
```

```{r fl.plot}
do.heatmap <- function(res) {
  fl <- res$fl
  
  FF <- ldf(fl, type = "I")$F
  FF <- FF[, -1]
  FF <- FF[, order(res$fl$pve[-1], decreasing = TRUE)]
  colnames(FF) <- 1:ncol(FF) 
  
  cell_type <- pbmc.celltype
  
  tib <- as_tibble(FF) %>%
    mutate(Cell.type = cell_type)
  
  tsne_res <- Rtsne(
    as.matrix(tib %>% select(-Cell.type)),
    dims = 1,
    perplexity = pmax(1, floor((nrow(tib) - 1) / 3) - 1),
    pca = FALSE,
    normalize = FALSE,
    theta = 0.1,
    check_duplicates = FALSE,
    verbose = FALSE
  )$Y[, 1]
  tib <- tib %>%
    mutate(tsne_res = unlist(tsne_res)) %>%
    arrange(Cell.type, tsne_res) %>%
    mutate(Cell.idx = row_number()) %>%
    select(-tsne_res)
  
  cell_type <- tib$Cell.type
  
  tib <- tib %>%
    pivot_longer(
      -c(Cell.idx, Cell.type),
      names_to = "Factor",
      values_to = "Loading",
      values_drop_na = TRUE
    ) %>%
    mutate(Factor = as.numeric(Factor))
  
  cell_type_breaks <- c(1, which(cell_type[2:nrow(tib)] != cell_type[1:(nrow(tib) - 1)]))
  ggplot(tib, aes(x = Factor, y = -Cell.idx, fill = Loading)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "red") +
    labs(y = "") +
    scale_y_continuous(breaks = -cell_type_breaks,
                       minor_breaks = NULL,
                       labels = unique(tib$Cell.type)) +
    theme_minimal() +
    geom_hline(yintercept = -cell_type_breaks, size = 0.1)
}
```

I give a semi-nonnegative and three non-negative `flashier` fits to the PBMCs dataset that I also looked at in my thesis. The nonnegative fits were obtained by backfitting from a greedy nonnegative fit and from a NNMF obtained via `NNLM`.

### Semi-nonnegative fit

```{r smnf, results = 'asis'}
smnf <- readRDS("./output/pbmc/smnf.rds")
cat("ELBO:", smnf$fl$elbo, "\n")
units(smnf$t) <- "mins"
cat("Time to fit:", format(smnf$t, digits = 1), "\n")
cat("Maximum iterations reached:", !is.null(smnf$fl$flash.fit$maxiter.reached), "\n")
do.heatmap(smnf)
```

<!-- ### Greedy nonnegative fit -->

<!-- ```{r greedy2} -->
<!-- greedy2 <- readRDS("./output/pbmc/greedy2.rds") -->
<!-- do.heatmap(greedy2) -->
<!-- ``` -->

### Nonnegative fit (from greedy)

```{r bf2, results = 'asis'}
greedy <- readRDS("./output/pbmc/greedy.rds")
bf <- readRDS("./output/pbmc/bf.rds")
bf.t <- bf$t + greedy$t
cat("ELBO:", bf$fl$elbo, "\n")
units(bf.t) <- "mins"
cat("Time to fit:", format(bf.t, digits = 1), "\n")
cat("Maximum iterations reached:", !is.null(bf$fl$flash.fit$maxiter.reached), "\n")
do.heatmap(bf)
```

<!-- ### NNLM fit -->

<!-- ```{r nnlm} -->
<!-- nnlm <- readRDS("./output/pbmc/nnlm.rds") -->
<!-- do.heatmap(nnlm) -->
<!-- ``` -->

### Nonnegative fit (from NNLM)

```{r nnlmbf, results = 'asis'}
nnlmbf <- readRDS("./output/pbmc/nnlmbf.rds")
cat("ELBO:", nnlmbf$fl$elbo, "\n")
units(nnlmbf$t) <- "mins"
cat("Time to fit:", format(nnlmbf$t, digits = 1), "\n")
cat("Maximum iterations reached:", !is.null(nnlmbf$fl$flash.fit$maxiter.reached), "\n")
do.heatmap(nnlmbf)
```

### Nonnegative fit (from NNLM, shifted point-exponential priors)
```{r nzpe, results = 'asis'}
nzpe <- readRDS("./output/pbmc/nzpe.rds")
nzpe2 <- readRDS("./output/pbmc/nzpe2.rds")
# shifts <- sapply(nzpe$fl$F.ghat[-1], function(k) k$shift[1])
# nzpe$fl$flash.fit$EF[[2]][, -1] <- nzpe$fl$flash.fit$EF[[2]][, -1] - rep(shifts, each = ncol(pbmc$fl.dat))
nzpe.t <- nzpe$t + nzpe2$t
cat("ELBO:", nzpe2$fl$elbo, "\n")
units(nzpe.t) <- "mins"
cat("Time to fit:", format(nzpe.t, digits = 1), "\n")
cat("Maximum iterations reached:", !is.null(nzpe2$fl$flash.fit$maxiter.reached), "\n")
do.heatmap(nzpe2)
```
