---
title: "Pijuan-Sala Tal1 depletion dataset: semi-nonnegative fit (preliminary)"
author: "Jason Willwerscheid"
date: "4/13/2022"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

```{r read.data}
library(tidyverse)
library(flashier)
library(ggrepel)
library(Matrix)
library(xtable)
library(gt)
```

The subset of endoderm and gut cells discussed in [Pijuan-Sala et al.](https://www.nature.com/articles/s41586-019-0933-9):

```{r fl.plot}
res <- readRDS("./output/pijuan/endoderm.snmf.small")

do.heatmap <- function(res, plot.group, rev = FALSE) {
  FF <- res$fit$F[, -1]
  colnames(FF) <- 1:ncol(FF) 
  
  tib <- as_tibble(FF) %>%
    mutate(Plot.group = plot.group) 
  
  pca_res <- prcomp(
    as.matrix(tib %>% select(-Plot.group)),
    retx = TRUE,
    center = TRUE,
    scale. = FALSE
  )$x[, 1]
  if (rev) {
    pca_res = -pca_res
  }
  tib <- tib %>%
    mutate(pca_res = pca_res) 

  tib <- tib %>%
    arrange(Plot.group, pca_res) %>%
    select(-pca_res) %>%
    mutate(Cell.idx = row_number())
        
  plt_grp <- tib$Plot.group

  tib <- tib %>%
    pivot_longer(
      -c(Cell.idx, Plot.group),
      names_to = "Factor",
      values_to = "Loading",
      values_drop_na = TRUE
    ) %>%
    mutate(Factor = as.numeric(Factor))
  
  grp_breaks <- c(1, which(plt_grp[2:nrow(tib)] != plt_grp[1:(nrow(tib) - 1)]))
  ggplot(tib, aes(x = Factor, y = -Cell.idx, fill = Loading)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "red") +
    labs(y = "") +
    scale_y_continuous(breaks = -grp_breaks,
                       minor_breaks = NULL,
                       labels = unique(tib$Plot.group)) +
    theme_minimal() +
    geom_hline(yintercept = -grp_breaks, size = 0.1)
}

do.heatmap(res, factor(res$celltype, levels = c(
  "Anterior Primitive Streak", "Def. endoderm", "Gut", "Visceral endoderm"
)), rev = TRUE)
```

* Factor 1: visceral endoderm -> gut
* Factor 2: maybe E7.25 or batch 2?
* Factor 3: anterior primitive streak -> gut
* Factor 4: mostly E7.25 to E7.75
* Factor 5: mixed gastrulation
* Factor 6: TS12 (and gut)
* Factor 7: TS11
* Factor 8: VE -> gut
* Factor 9: mostly E7.75 to E8.25
* Factor 10: ??
* Factor 11: foregut 1, midgut, VE subpopulation
* Factor 12: VE subpopulation
* Factor 13: hindgut
* Factors 14/15: DE subpopulations



```{r}
top.genes <- apply(res$fit$L[, -1], 2, function(k) {
  res$genelist[order(k, decreasing = TRUE)[1:12]]
})
colnames(top.genes) <- paste0("k=", formatC(1:ncol(top.genes), width = 2, flag = "0"))

genes.tib <- as_tibble(top.genes) %>%
  pivot_longer(cols = everything(), names_to = "Factor", values_to = "SYMBOL")

genes.tib <- genes.tib %>%
  group_by(Factor) %>%
  summarize(TopGenes = paste(SYMBOL, collapse = ", "))

tbl <- genes.tib %>%
  gt() %>%
  cols_label(Factor = "Factor", TopGenes = "Top Genes (pos. loadings only)") %>%
  cols_align("left", columns = Factor) %>%
  cols_align("left", columns = TopGenes) %>%
  cols_width(
    Factor ~ pct(10),
    TopGenes ~ pct(90)
  ) %>%
  opt_row_striping()

tbl
```

```{r more.heatmaps}
do.heatmap(res, factor(res$theiler, levels = c(
  "TS9", "TS9-10", "TS10", "TS11", "TS12"
)))

do.heatmap(res, factor(res$stage, levels = c(
  "E6.5", "E6.75", "E7.0", "E7.25", "E7.5", "E7.75", "E8.0", "E8.25", "E8.5", "mixed_gastrulation"
)))

do.heatmap(res, factor(res$batch))

cell.type <- ifelse(is.na(res$gut.type), res$celltype, res$gut.type)
cell.type <- factor(cell.type, levels = c(
  "Anterior Primitive Streak", "Def. endoderm", "Pharyngeal endoderm", "Foregut 2", "Foregut 1",
  "Midgut", "Midgut/Hindgut", "Hindgut 2", "Hindgut 1", "Gut", "Visceral endoderm"
))
do.heatmap(res, cell.type)
```

```{r volcano.code}
# do.volcano.plot <- function(res, k, plt.title = "") {
#   genes <- res$genelist
#   fl <- res$fit
#   
#   k <- order(fl$pve[-1], decreasing = TRUE)[k] + 1
#   
#   tib <- tibble(
#     pm = res$fit$L,
#     z = abs(res$fit$L.pm[, k]) / pmax(sqrt(.Machine$double.eps), res$fit$L.psd[, k]),
#     SYMBOL = genes
#   ) %>%
#     mutate(SYMBOL = ifelse(
#       z > sort(z, decreasing = TRUE)[11] |
#         (pm > sort(pm, decreasing = TRUE)[11]) |
#         (pm < sort(pm, decreasing = FALSE)[11]), 
#       SYMBOL, 
#       ""
#     )) 
#   
#   plt <- ggplot(tib, aes(x = pm, y = z, label = SYMBOL)) +
#     geom_point() +
#     # scale_color_gradient2(low = "deepskyblue", mid = "gold", high = "orangered",
#     #                       na.value = "gainsboro",
#     #                       midpoint = mean(range(tib$exprmean))) +
#     scale_y_sqrt() +
#     geom_text_repel(color = "darkgray",size = 2.25, fontface = "italic",
#                     segment.color = "darkgray", segment.size = 0.25,
#                     min.segment.length = 0, na.rm = TRUE, max.overlaps = 20) +
#     theme_minimal() +
#     labs(
#       x = "Factor Loading (posterior mean)",
#       y = "|z-score|",
#       # color = "Mean Expression (log10)",
#       title = plt.title
#     ) +
#     theme(legend.position = "bottom")
# 
#   return(plt)
# }
# 
# do.volcano.plot(res, 1, "VE (1)")
# do.volcano.plot(res, 3, "PS -> Gut (3)")
# do.volcano.plot(res, 6, "Gut (6)")
# do.volcano.plot(res, 8, "VE -> Gut (8)")
# do.volcano.plot(res, 11, "Midgut (11)")
# do.volcano.plot(res, 12, "VE subpop (12)")
# do.volcano.plot(res, 13, "Hindgut (13)")
# do.volcano.plot(res, 14, "DE subpop (14)")
# do.volcano.plot(res, 15, "DE subpop (15)")
```
