---
title: "Prior families"
author: "Jason Willwerscheid"
date: "3/18/2020"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

I redo my [previous analysis of prior families](prior_type.html) using the PBMC 3k dataset. Fits were produced by adding 20 "greedy" factors and backfitting. The code can be viewed [here](https://github.com/willwerscheid/scFLASH/blob/master/code/prior_type/priortype_fits_pbmc.R).

```{r load.res}
source("./code/utils.R")
pbmc <- readRDS("./data/10x/pbmc.rds")
pbmc <- preprocess.pbmc(pbmc)
res <- readRDS("./output/prior_type/priortype_fits_pbmc.rds")
```

## Results: Fitting time

```{r fit.time}
format.t <- function(t) {
  return(sapply(t, function(x) {
    hrs <- floor(x / 3600)
    min <- floor((x - 3600 * hrs) / 60)
    sec <- floor(x - 3600 * hrs - 60 * min)
    if (hrs > 0) {
      return(sprintf("%02.fh%02.fm%02.fs", hrs, min, sec))
    } else if (min > 0) {
      return(sprintf("%02.fm%02.fs", min, sec))
    } else {
      return(sprintf("%02.fs", sec))
    }
  }))
}

t.greedy <- sapply(lapply(res, `[[`, "t"), `[[`, "greedy")
t.backfit <- sapply(lapply(res, `[[`, "t"), `[[`, "backfit")
niter.backfit <- sapply(lapply(res, `[[`, "output"), function(x) x$Iter[nrow(x)])
t.periter.b <- t.backfit / niter.backfit
time.df <- data.frame(format.t(t.greedy), format.t(t.backfit),
                      niter.backfit, format.t(t.periter.b))
rownames(time.df) <- c("Point-normal", "Scale mixture of normals",
                       "Semi-nonnegative (NN cells)", "Semi-nonnegative (NN genes)")
knitr::kable(time.df[c(1, 2, 4, 3), ],
             col.names = c("Greedy time", "Backfit time",
                           "Backfit iter", "Backfit time per iter"),
             digits = 2,
             align = "r")
```

## Results: ELBO

I show the ELBO after each of the last ten greedy factors have been added and after each backfitting iteration. `snn.cell` denotes the semi-nonnegative fit that puts nonnegative priors on cell loadings, whereas `snn.gene` puts nonnegative priors on gene loadings.

Results here are surprising: the `snn.gene` fit easily achieves the highest ELBO, and both semi-nonnegative fits outperform the point-normal fit. 

```{r pn.v.ash.elbo}
res$pn$output$Fit <- "pn"
res$ash$output$Fit <- "ash"
res$snn.cell$output$Fit <- "snn.cell"
res$snn.gene$output$Fit <- "snn.gene"

res$pn$output$row <- 1:nrow(res$pn$output)
res$ash$output$row <- 1:nrow(res$ash$output)
res$snn.cell$output$row <- 1:nrow(res$snn.cell$output)
res$snn.gene$output$row <- 1:nrow(res$snn.gene$output)

elbo.df <- rbind(res$pn$output, res$ash$output, res$snn.cell$output, res$snn.gene$output)
ggplot(subset(elbo.df, !(Factor %in% as.character(1:10))), 
       aes(x = row, y = Obj, color = Fit, shape = Type)) +
  geom_point() +
  labs(x = NULL, y = "ELBO (unadjusted)",
       title = "Greedy factors 11-20 and all backfitting iterations") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

## Results: Point-normal priors vs. scale mixtures of normals

I line up the point-normal and scale-mixture-of-normal factors so that similar factors are shown one on top of the other. Although many factors appear similar, the scale-mixture-of-normal fit looks much better to me: factor 6 does a better job at separating out dendritic cells, while factor 14 contains potentially useful information about CD19+ B cells.

```{r pn.v.ash.factors}
pn.v.ash <- compare.factors(res$pn$fl, res$ash$fl)
plot.factors(res$pn, pbmc$cell.type, order(res$pn$fl$pve, decreasing = TRUE), 
             title = "Point-normal priors")
plot.factors(res$ash, pbmc$cell.type, pn.v.ash$fl2.k[order(res$pn$fl$pve, decreasing = TRUE)],
             title = "Scale mixtures of normals")
```

## Results: Semi-nonnegative factorizations

For ease of comparison, I only consider the semi-nonnegative fit that puts nonnegative priors on gene loadings. As above, I line up similar factors. I denote a semi-nonnegative factor as "similar" to a point-normal factor if the gene loadings are strongly correlated ($r \ge 0.7$) with either the positive or the negative component of the point-normal gene loadings. 

```{r snn.matched}
cor.thresh <- 0.7

pn.pos <- pmax(res$ash$fl$loadings.pm[[2]], 0)
pn.pos <- t(t(pn.pos) / apply(pn.pos, 2, function(x) sqrt(sum(x^2))))
pos.cor <- crossprod(res$snn.gene$fl$loadings.pm[[2]], pn.pos)

pn.neg <- -pmin(res$ash$fl$loadings.pm[[2]], 0)
pn.neg <- t(t(pn.neg) / apply(pn.neg, 2, function(x) sqrt(sum(x^2))))
pn.neg[, 1] <- 0
neg.cor <- crossprod(res$snn.gene$fl$loadings.pm[[2]], pn.neg)

is.cor <- (pmax(pos.cor, neg.cor) > cor.thresh)

pn.matched <- which(apply(is.cor, 2, any))
snn.matched <- unlist(lapply(pn.matched, function(x) which(is.cor[, x])))
# Duplicate factors where need be.
pn.matched <- rep(1:res$ash$fl$n.factors, times = apply(is.cor, 2, sum))

plot.factors(res$ash, pbmc$cell.type,
             pn.matched, title = "Point-normal (matched factors)")
plot.factors(res$snn.gene, pbmc$cell.type,
             snn.matched, title = "Semi-nonnegative (matched factors)")
```

The remaining factors do not have counterparts. 

```{r snn.unmatched}
snn.unmatched <- setdiff(1:res$snn.gene$fl$n.factors, snn.matched)
pn.unmatched <- setdiff(1:res$ash$fl$n.factors, pn.matched)

plot.factors(res$ash, pbmc$cell.type,
             pn.unmatched, title = "Point-normal (unmatched factors)")
plot.factors(res$snn.gene, pbmc$cell.type,
             snn.unmatched, title = "Semi-nonnegative (unmatched factors)")
```
