---
title: "Pseudocounts"
author: "Jason Willwerscheid"
date: "3/19/2020"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

I redo my [previous analysis of pseudocounts](pseudocount_redux.html) using the PBMC 3k dataset. Fits were produced by adding 20 "greedy" factors and backfitting. The code can be viewed [here](https://github.com/willwerscheid/scFLASH/blob/master/code/pseudocount_redux/pseudocount_redux_pbmc.R).

```{r load.res}
source("./code/utils.R")
pbmc <- readRDS("./data/10x/pbmc.rds")
pbmc <- preprocess.pbmc(pbmc)
res <- readRDS("./output/pseudocount/pseudocount_fits_pbmc.rds")
```

## Results: ELBO

As expected, the ELBO is monotonically decreasing as a function of the pseudocount.

```{r elbo}
elbo.df <- data.frame(pseudocount = as.numeric(names(res)),
                      elbo = sapply(res, function(x) x$fl$elbo + x$elbo.adj))
ggplot(elbo.df, aes(x = pseudocount, y = elbo)) +
  geom_point() +
  scale_x_continuous(trans = "log2") +
  labs(y = "ELBO")
```

## Results: Log likelihood of implied distribution

An in the previous analysis, $\alpha = 0.5$ does best.

```{r llik}
llik.df <- data.frame(pseudocount = as.numeric(names(res)),
                      llik = sapply(res, function(x) x$p.vals$llik))
ggplot(llik.df, aes(x = pseudocount, y = llik)) +
  geom_point() +
  scale_x_continuous(trans = "log2") +
  labs(y = "log likelihood (implied model)")
```

## Results: p-values

```{r p.vals, results = "asis"}
for (pc in names(res)) {
  cat("\n### Pseudocount = ", pc, "\n")
  plot(plot.p.vals(res[[pc]][["p.vals"]]))
  cat("\n")
}
```

