---
title: "SNMF on Data Simulated from an NMF model"
author: "Jason Willwerscheid"
date: "3/24/2022"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

```{r sim_fn}
library(flashier)
library(tidyverse)

sim_data <- function(n = 100, 
                     p = 200, 
                     K = 6, 
                     L.nn = 10, 
                     F.nn = 20, 
                     se = 0.1, 
                     K.dense = 1, 
                     seed = 666) {
  set.seed(seed)
  
  LL <- matrix(rexp(n * K), nrow = n, ncol = K)
  FF <- matrix(rexp(p * K), nrow = p, ncol = K)
  
  # "Mean" factor.
  LL[, 1] <- 3
  
  # Additional sparse nonnegative factors.
  for (k in (K.dense + 1):K) {
    L.nn.idx <- seq((k - K.dense - 1) * L.nn + 1, (k - K.dense) * L.nn)
    F.nn.idx <- seq((k - K.dense - 1) * (F.nn / 2) + 1, (k + K.dense) * (F.nn / 2))
    LL[setdiff(1:n, L.nn.idx), k] <- 0
    FF[setdiff(1:p, F.nn.idx), k] <- 0
  }
  
  # Add normal noise.
  Y <- LL %*% t(FF) + rnorm(n * p, sd = se)
  
  # Add a constant (which can be absorbed by mean factor) to ensure nonnegativity.
  Y <- Y - min(Y)
  
  return(list(LL = LL, FF = FF, Y = Y))
}

plot_it <- function(simdat, nnmf_res, snmf_res) {
  LL <- simdat$LL

  to_tibble <- function(mat, type) {
    mat <- scale(mat, center = FALSE, scale = apply(mat, 2, function(x) max(abs(x))))
    return(
      as_tibble(mat, .name_repair = "unique") %>%
        mutate(row = row_number()) %>%
        pivot_longer(!row, names_to = "k", values_to = "value") %>%
        add_column(type = type)     
    )
  }
  
  
  suppressMessages({
    tib <- to_tibble(simdat$LL, "True loadings") %>%
      bind_rows(to_tibble(simdat$FF, "True factors")) %>%
      bind_rows(to_tibble(snmf_res$L.pm, "SNMF loadings")) %>%
      bind_rows(to_tibble(snmf_res$F.pm, "SNMF factors")) 
    if (inherits(nnmf_res, "flash")) {
      tib <- tib %>%
        bind_rows(to_tibble(nnmf_res$L.pm, "NMF loadings")) %>%
        bind_rows(to_tibble(nnmf_res$F.pm, "NMF factors")) 
    } else {
      tib <- tib %>%
        bind_rows(to_tibble(nnmf_res$W, "NMF loadings")) %>%
        bind_rows(to_tibble(t(nnmf_res$H), "NMF factors")) 
    }
    tib <- tib %>%
      mutate(k = as.numeric(str_remove_all(k, "\\."))) %>%
      mutate(type = factor(type, levels = c(
        "True loadings", "NMF loadings", "SNMF loadings",
        "True factors", "NMF factors", "SNMF factors"
      )))
  })
  
  ggplot(tib, aes(x = k, y = row, fill = value)) +
    geom_tile() +
    scale_fill_gradient2() +
    facet_wrap(~type, nrow = 2, ncol = 3, dir = "h", scales = "free") +
    theme_void()
 }
```

I simulate data 
$$ Y = LF' + E $$ 
where both $L$ and $F$ are nonnegative and $E$ is Gaussian noise. For details, refer to the code above. I refer to this model as the NMF model. I then fit an NMF using `NNLM` with the correct number of factors, as well as semi-nonnegative EBMF using `flashier` with the correct number of factors:
```{r sim1_trueK}
sim1 <- sim_data(K = 6)

# Use true value of K.
nnmf_res1 <- NNLM::nnmf(sim1$Y, k = 6, verbose = 0)
snmf_res1 <- flash.init(sim1$Y) %>%
  flash.set.verbose(0) %>%
  flash.add.greedy(
    Kmax = 6,
    ebnm.fn = c(ebnm::ebnm_point_normal, ebnm::ebnm_point_exponential),
    init.fn = function(fl) init.fn.default(fl, dim.signs = c(0, 1))
  ) %>%
  flash.backfit() %>%
  flash.nullcheck()

plot_it(sim1, nnmf_res1, snmf_res1)
```

Not only is SNMF basically nonnegative, but it does a better job recovering the true factors than NMF!

Next I fit NMF and SNMF using more factors than necessary:

```{r sim1_nottrueK}
nnmf_res1b <- NNLM::nnmf(sim1$Y, k = 10, verbose = 0)
snmf_res1b <- flash.init(sim1$Y) %>%
  flash.set.verbose(0) %>%
  flash.add.greedy(
    Kmax = 10,
    ebnm.fn = c(ebnm::ebnm_point_normal, ebnm::ebnm_point_exponential),
    init.fn = function(fl) init.fn.default(fl, dim.signs = c(0, 1))
  ) %>%
  flash.backfit() %>%
  flash.nullcheck()

plot_it(sim1, nnmf_res1b, snmf_res1b)
```

SNMF (almost) recovers the correct number of factors, which NMF is of course unable to do. Interestingly, however, if I use EBMF to fit NMF, I get approximately the same fit as SNMF:

```{r sim1_nnebmf}
nnmf_res1c <- flash.init(sim1$Y) %>%
  flash.set.verbose(0) %>%
  flash.add.greedy(
    Kmax = 10,
    ebnm.fn = ebnm::ebnm_point_exponential,
    init.fn = function(fl) init.fn.default(fl, dim.signs = c(1, 1))
  ) %>%
  flash.backfit() %>%
  flash.nullcheck()

plot_it(sim1, nnmf_res1c, snmf_res1b)
```

Finally I make the simulation scenario a bit more difficult (more noise, less sparsity, additional dense factors) and try again with the correct number of factors:

```{r sim2}
sim2 <- sim_data(K = 8, se = 1, L.nn = 20, F.nn = 40, K.dense = 3)

nnmf_res2 <- NNLM::nnmf(sim2$Y, k = 8, verbose = 0)
snmf_res2 <- flash.init(sim2$Y) %>%
  flash.set.verbose(0) %>%
  flash.add.greedy(
    Kmax = 8,
    ebnm.fn = c(ebnm::ebnm_point_normal, ebnm::ebnm_point_exponential),
    init.fn = function(fl) init.fn.default(fl, dim.signs = c(0, 1))
  ) %>%
  flash.backfit() %>%
  flash.nullcheck()

plot_it(sim2, nnmf_res2, snmf_res2)
```

SNMF again does quite a bit better on the sparse factors, but the dense factors now show a mixture of positive and negative loadings.

I try this more complex scenario again but use EBMF to fit NMF:

```{r sim2b}
nnmf_res2b <- flash.init(sim2$Y) %>%
  flash.set.verbose(0) %>%
  flash.add.greedy(
    Kmax = 8,
    ebnm.fn = ebnm::ebnm_point_exponential,
    init.fn = function(fl) init.fn.default(fl, dim.signs = c(1, 1))
  ) %>%
  flash.backfit() %>%
  flash.nullcheck()

plot_it(sim2, nnmf_res2b, snmf_res2)
```

