---
title: "Variance Regularization: Part One"
author: "Jason Willwerscheid"
date: "3/10/2020"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Although I prefer a gene-wise variance structure, it can cause problems for backfits. When the low-rank EBMF structure very closely approximates the observed expression for a particular gene, then the estimated residual variance for that gene can blow up to infinity. This typically happens with very sparsely expressed genes, so removing genes with zero counts in a large majority of cells mitigates the problem. I don't want to remove too many genes, however, since sparsely expressed genes can provide important information about rare cell types (e.g., Cftr and ionocytes).

Up until now, I've dealt with this problem is a very ad hoc manner: I set the minimum gene-wise residual variance obtained from the greedy fit as the minimum for all residual variances estimated during the backfit. Here I explore a less ad-hoc approach in which I put a prior on the gene-wise precisions:

$$ 1 / \sigma_j^2 \sim \text{Exponential}(\lambda) $$

The $\sigma_j$s can then be estimated by solving an empirical Bayes Poisson means (EBPM) problem. The `ebpm` package is not yet able to solve this particular EBPM problem, so in this analysis I'll fix $\lambda$ and explore results for various choices. (I attempted to use `exponential_mixture` and `point_gamma` prior families, but they did a very poor job at regularizing the variance estimates.)

All fits "pre-scale" cells, add 20 factors greedily using point-normal priors, and then backfit. The code used to produce the fits can be viewed [here](https://github.com/willwerscheid/scFLASH/blob/master/code/backfit/backfit_fits.R).

```{r load.res}
suppressMessages(library(tidyverse))
suppressMessages(library(Matrix))

source("./code/utils.R")
pbmc <- readRDS("./data/10x/pbmc.rds")
pbmc <- preprocess.pbmc(pbmc)

res <- readRDS("./output/var_reg/varreg_fits.rds")
```

## Results: ELBO

As the scale of the prior increases, less shrinkage will be applied, so that estimates will be closer to the MLE estimates and a larger ELBO will result. Interestingly, the same is not true for the likelihood of the implied discrete distribution, which peaks somewhere around $\lambda = 100$ or $\lambda = 1000$.

This order of magnitude for $\lambda$ makes intuitive sense. I've retained genes that have nonzero counts in at least 10 of the 3206 cells. If we regard the true marginal distribution for gene $j$ as $\text{Poisson}(\nu)$, then (using the ML estimate for $\nu$) its residual variance will be at least $10 / 3206$. (One can also show that if the true distribution is a mixture $\pi_1 \text{Poisson}(\nu_1) + \pi_2 \text{Poisson}(\nu_2)$, then the variance is bounded below by $\mu(1 - \mu)$, where $\mu$ is the mean of the distribution: $\mu = \pi_1 \nu_1 + \pi_2 \nu_2$. Plugging in $\hat{\mu} = 10/3206$ gives an upper bound of 321.6 for gene-wise precisions.)

```{r progress}
prior.scale <- 10^(0:5)
elbo.df <- tibble(scale = prior.scale,
                  elbo = sapply(res, function(x) x$fl$elbo),
                  llik = sapply(res, function(x) x$p.vals$llik))

ggplot(elbo.df, aes(x = scale, y = elbo)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "Scale of prior",
       y = NULL,
       title = "ELBO")

ggplot(elbo.df, aes(x = scale, y = llik)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "Scale of prior",
       y = NULL,
       title = "Likelihood of implied discrete distribution")
```

## Results: Factor comparisons

Plotted factors look fairly similar for all choices of $\lambda$, especially for more reasonable choices ($\lambda > 100$). 

```{r factors}
plot.factors(res$scale1, pbmc$cell.type, kset = order(res$scale1$fl$pve, decreasing = TRUE),
             title = "Scale = 1")
plot.factors(res$scale100, pbmc$cell.type, kset = order(res$scale100$fl$pve, decreasing = TRUE), 
             title = "Scale = 100")
plot.factors(res$scale10000, pbmc$cell.type, kset = order(res$scale10000$fl$pve, decreasing = TRUE),
             title = "Scale = 10000")
```

